Paper_Figure

### 2023-10-18 Marker_identfication_benchmark.Rmd
robustness test

# Load Data & Process
## Load Data
```{r}
path <- "/data/ruiqi/local_marker/dataset/tabular_muris"
tiss <- readRDS(paste0(path,"/marrow_facs.rds"))
```
## Process
```{r}
DefaultAssay(tiss) <- "RNA"
n_dim = dim(tiss@reductions$pca@cell.embeddings)[2]
feature_space = as.matrix(tiss@reductions$pca@cell.embeddings[,1:n_dim])
visual_space = data.frame(tiss@reductions$tsne@cell.embeddings)
dat = as.matrix(tiss[[DefaultAssay(tiss)]]@data)
cell_label = tiss$cell_ontology_class

@@

# Define Ground truth Marker
## Criterion1: Fold-change
For each gene, sorting cell types by mean expression for each gene and computing fold changes between consecutive types, take the maximum value among the N-1 fold change values, given N cell types.
```{r}
path = "./benchmark_result/"
tissue_name = "marrow_facs"
```

@@

# Run Each Method and save results
```{r}
path = "./benchmark_result/"
tissue_name = "marrow_facs"
tissue_name = "lung_facs"
```

## LocalMarkerDetector
Genes were ranked by the increasing order of Cumulative Diffusion-KL score
```{r}
method = "lmd"
source("./LMD_function.R", echo = F)
res = LMD(dat,feature_space,max_time = 2^10)
marker = show_result_lmd(res, n = length(selected_genes))
marker[order(marker$score),'rank'] = 1:nrow(marker)
write.table(marker,file = sprintf("%s%s_%s.csv",path,method,tissue_name),row.names = TRUE)
```

@@

# Integrate rank from all methods into one dataframe
```{r}
path = "./benchmark_result/"
tissue_name = "marrow_facs"
```
```{r}
method_ls = c("lmd","haystack","hotspot","marcopolo","semi","wilcox_default","wilcox_no_filter","hvg")
method_fullname = setNames(c("LMD","singleCellHaystack","Hotspot","Marcopolo","SEMITONES","Seurat v3 default", "Seurat v3 Wilcoxon unfiltered","HighlyVariableGenes"),method_ls)

@@

# AUROC
```{r}
path = "./benchmark_result/"
tissue_name = "marrow_facs"
df_benchmark = read.table(paste0(path,tissue_name,"_benchmark_rank_table.csv"))
```
```{r}
max_logfc = read.table(paste0(path,tissue_name,"_ground_truth_c1.txt")) %>% rownames()
celldb_marker = read.table(paste0(path,tissue_name,"_ground_truth_c2.txt"))[,1]

@@

# Density Index
```{r}
path = "./benchmark_result/"
tissue_name = "marrow_facs"
```
```{r}
df_benchmark = read.table(paste0(path,tissue_name,"_benchmark_rank_table.csv"))

@@

# Sensitivity Test (Try different K)
```{r}
path = "./marrow_result/"
tissue_name = "marrow_facs"
```
```{r}
# RunLMD under differen K
score_df = do.call(cbind,lapply(c(3:10,20,30,40,50),function(kNN){
  try({
    res = LMD(dat,feature_space,knn = kNN)
    return(res$cumulative_score)
  },silent = TRUE)
}))
colnames(score_df) = paste0("kNN",c(3:10,20,30,40,50))
write.table(score_df,file = paste0(path,tissue_name,"_different_knn.csv"),row.names = FALSE)
```
```{r}
# Obtain AUROC
score_ranked <- data.frame(score_df) %>% mutate(across(everything(), rank))

max_logfc = read.table(paste0(path,tissue_name,"_ground_truth_c1.txt")) %>% rownames()
celldb_marker = read.table(paste0(path,tissue_name,"_ground_truth_c2.txt"))[,1]

@@

auc_df = apply(score_ranked,2,function(x){
  unlist(lapply(gt_genes_list,function(genes){
    roc = pROC::roc(as.integer(rownames(score_ranked) %in% genes), x, direction = ">")
    pROC::auc(roc)
  }))
})
auc_df = data.frame(auc_df,gt = names(gt_list))
write.table(auc_df,file = paste0(path,tissue_name,"_sensitivity_auroc.csv"),row.names = FALSE)
```

>>>>

# Load Data & Process
## Load Data
```{r}
path <- "/data/ruiqi/local_marker/dataset/tabular_muris/"
tissue_name = "marrow_facs" #lung_facs, pancreas_facs
tiss <- readRDS(paste0(path,tissue_name,".rds"))
```
## Process
```{r}
DefaultAssay(tiss) <- "RNA"
n_dim = dim(tiss@reductions$pca@cell.embeddings)[2]
feature_space = as.matrix(tiss@reductions$pca@cell.embeddings[,1:n_dim])
visual_space = data.frame(tiss@reductions$tsne@cell.embeddings)
dat = as.matrix(tiss[[DefaultAssay(tiss)]]@data)
cell_label = tiss$cell_ontology_class

@@

# Define Ground truth Marker
## Criterion1: Fold-change
For each gene, sorting cell types by mean expression for each gene and computing fold changes between consecutive types, take the maximum value among the N-1 fold change values, given N cell types.
```{r}
path = "./benchmark_result/"
# tissue_name = "marrow_facs"
```

@@

# Run Each Method and save results
```{r}
path = "./benchmark_result/"
# tissue_name = "marrow_facs"
```

## LocalMarkerDetector
Genes were ranked by the increasing order of Cumulative Diffusion-KL score
```{r}
method = "lmd"
source("./LMD_function.R", echo = F)
res = LMD(dat,feature_space,max_time = 2^10,knn = 7)
marker = show_result_lmd(res, n = length(selected_genes))
marker[order(marker$score),'rank'] = 1:nrow(marker)
write.table(marker,file = sprintf("%s%s_%s.csv",path,method,tissue_name),row.names = TRUE)
```

@@

# Integrate rank from all methods into one dataframe
```{r}
path = "./benchmark_result/"
# tissue_name = "marrow_facs"
```
```{r}
method_ls = c("lmd","haystack","hotspot","marcopolo","semi","wilcox_default","wilcox_no_filter","hvg")
method_fullname = setNames(c("LMD","singleCellHaystack","Hotspot","Marcopolo","SEMITONES","Seurat v3 default", "Seurat v3 Wilcoxon unfiltered","HighlyVariableGenes"),method_ls)

@@

# AUROC
```{r}
path = "./benchmark_result/"
# tissue_name = "marrow_facs"
df_benchmark = read.table(paste0(path,tissue_name,"_benchmark_rank_table.csv"))
```
```{r}
max_logfc = read.table(paste0(path,tissue_name,"_ground_truth_c1.txt")) %>% rownames()
celldb_marker = read.table(paste0(path,tissue_name,"_ground_truth_c2.txt"))[,1]

@@

# Density Index
```{r}
path = "./benchmark_result/"
# tissue_name = "marrow_facs"
```
```{r}
df_benchmark = read.table(paste0(path,tissue_name,"_benchmark_rank_table.csv"))

@@

# Sensitivity Test (Try different K)
```{r}
path = "./marrow_result/"
# tissue_name = "marrow_facs"
```
```{r}
# RunLMD under differen K
source("./LMD_function.R", echo = F)
score_df = do.call(cbind,lapply(c(3:10,20,30,40,50),function(kNN){
  try({
    res = LMD(dat,feature_space,knn = kNN)
    return(res$cumulative_score)
  },silent = TRUE)
}))
colnames(score_df) = paste0("kNN",c(3:10,20,30,40,50))
write.table(score_df,file = paste0(path,tissue_name,"_different_knn.csv"),row.names = FALSE)
```
```{r}
# Obtain AUROC
score_ranked <- data.frame(score_df) %>% mutate(across(everything(), rank))
path = "./benchmark_result/"
max_logfc = read.table(paste0(path,tissue_name,"_ground_truth_c1.txt")) %>% rownames()
celldb_marker = read.table(paste0(path,tissue_name,"_ground_truth_c2.txt"))[,1]

@@

auc_df = apply(score_ranked,2,function(x){
  unlist(lapply(gt_list,function(genes){
    roc = pROC::roc(as.integer(rownames(score_ranked) %in% genes), x, direction = ">")
    pROC::auc(roc)
  }))
})
path = "./marrow_result/"
auc_df = data.frame(auc_df,gt = names(gt_list))
write.table(auc_df,file = paste0(path,tissue_name,"_sensitivity_auroc.csv"),row.names = FALSE)
```

### 2023-10-18 paper_figure.Rmd
robustness test

# Create the plot
fig <- ggplot(df_auc, aes(x = gt_set, y = AUROC, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(AUROC, 2)), position = position_dodge(width = 0.85), vjust = -0.25, size = 2) +
  theme_minimal() +
  labs(x = "Ground-Truth Genesets", y = "AUROC", fill = "Method") + coord_cartesian(ylim = c(0.5, 1)) +
  theme(legend.position = "bottom")  +  scale_fill_manual(values= coldef[names(coldef)%in%df_auc$Method])
fig
```

@@

# Fig - Sensitivity Test
```{r}
## Rank Stability
score_df = read.table(paste0(path,tissue_name,"_different_knn.csv"),header = TRUE)
score_ranked <- data.frame(score_df) %>% mutate(across(everything(), rank))
base = "kNN5"
top_genes_list <- lapply(c(50,seq(100,1000,100)), 
                         function(x) head(score_ranked[order(score_ranked[,base]),], x))
average_ranks <- lapply(top_genes_list, function(top_genes) {
  colMeans(top_genes)
})
average_ranks <- data.frame(
  Top_Genes = c(50,seq(100,1000,100)),
  do.call(rbind, average_ranks)
)
average_ranks <- average_ranks %>% tidyr::pivot_longer(cols = -Top_Genes, names_to = "kNN", values_to = "Average_Rank")
average_ranks$kNN = as.numeric(gsub("kNN","",average_ranks$kNN))
fig = ggplot(data = average_ranks, aes(x = kNN, y = Average_Rank, colour = factor(Top_Genes), group = Top_Genes)) +
  geom_line() +
  geom_point() +
  xlab("Top Genes") +
  ylab(sprintf("Average Rank of Top # Genes\n (base on %s)",base)) +
  theme_minimal() + facet_wrap(~Top_Genes,scales = "free")
fig
```
```{r}
## AUC stability
gt_genes_list <- c(lapply(c(50,seq(100,400,100)), function(x){
  names(max_logfc)[1:x]
}),list(true_marker))
auc_df = apply(score_ranked,2,function(x){
  unlist(lapply(gt_genes_list,function(genes){
    roc = pROC::roc(as.integer(rownames(score_ranked) %in% genes), x, direction = ">")
    pROC::auc(roc)
  }))
})
auc_df = data.frame(auc_df,gt = c(paste0("top",c(50,seq(100,400,100))),"cellMarkerDB"))
auc_df <- auc_df %>% tidyr::pivot_longer(cols = -gt, names_to = "kNN", values_to = "AUC")
auc_df$kNN = as.numeric(gsub("kNN","",auc_df$kNN))
ggplot(data = auc_df, aes(x = kNN, y = AUC, colour = factor(gt,levels = unique(gt)), group = gt)) +
  geom_line() +
  geom_point() +
  xlab("kNN") +
  ylab("AUC") +
  theme_minimal()

```

>>>>

# Create the plot
fig <- ggplot(df_auc, aes(x = gt_set, y = AUROC, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(AUROC, 2)), position = position_dodge(width = 0.85), vjust = -0.25, size = 2) +
  theme_minimal() +
  labs(x = "Ground-Truth Genesets", y = "AUROC", fill = "Method") + coord_cartesian(ylim = c(0.5, 1)) +
  theme(legend.position = "bottom")  +  scale_fill_manual(values= coldef[names(coldef)%in%df_auc$Method])
fig + ggtitle("Lung")
```

@@

# Fig - Sensitivity Test
```{r}
## Rank Stability
score_df = read.table(paste0(path,tissue_name,"_different_knn.csv"),header = TRUE)
score_ranked <- data.frame(score_df) %>% mutate(across(everything(), rank))
base = "kNN5"
top_genes_list <- lapply(c(50,seq(100,1000,100)), 
                         function(x) head(score_ranked[order(score_ranked[,base]),], x))
average_ranks <- lapply(top_genes_list, function(top_genes) {
  colMeans(top_genes)
})
average_ranks <- data.frame(
  Top_Genes = c(50,seq(100,1000,100)),
  do.call(rbind, average_ranks)
)
average_ranks <- average_ranks %>% tidyr::pivot_longer(cols = -Top_Genes, names_to = "kNN", values_to = "Average_Rank")
average_ranks$kNN = as.numeric(gsub("kNN","",average_ranks$kNN))

fig = ggplot(data = average_ranks, aes(x = kNN, y = Average_Rank, group = Top_Genes)) +
  geom_line() +
  geom_point() + labs(x = "kNN", y = sprintf("Average Rank of Top # Genes\n (base on %s)",base)) + geom_vline(xintercept = 5, color="red", linetype="dashed") +
  theme_minimal() + facet_wrap(~Top_Genes,scales = "free",labeller = as_labeller(setNames(paste0("Top",unique(average_ranks$Top_Genes)),unique(average_ranks$Top_Genes)))) + NoLegend()
pdf(width = 8,height = 5)
fig
dev.off()
```
```{r}
## AUC stability
auc_df = read.table(paste0(path,tissue_name,"_sensitivity_auroc.csv"),header = TRUE)
auc_df <- auc_df %>% tidyr::pivot_longer(cols = -gt, names_to = "kNN", values_to = "AUC") %>% filter(gt %in% c("Top100","CellMarkerDB"))
auc_df$gt = factor(auc_df$gt); levels(auc_df$gt) = c("CellMarkerDB","MacFC(Top100)")
auc_df$kNN = as.numeric(gsub("kNN","",auc_df$kNN))
fig = ggplot(data = auc_df, aes(x = kNN, y = AUC), group = gt) +
  geom_line() +
  geom_point() + labs(x = "kNN", y = "AUROC") + facet_wrap(~gt, scales = "free") + 
  theme_minimal()
pdf(width = 6,height = 3)
fig
dev.off()
```
